---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Tsion"
date: "02/26/26"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(mosaicData) 
```

### Instructional staff employment trends 

```{r read data and glimpse}
staff <- read_csv("data/instructional-staff.csv")
glimpse(staff)
```
Each row is a faculty type, and each year is stored in a separate column.

### Exercise 1

```{r change from wide to long}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))
staff_long
```


```{r Make a basic line graph of faculty type over time}
staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line(linewidth = 1) +
  labs(
    title = "Instructional Staff Employment Trends",
    x = "Year",
    y = "Percentage of instructional staff",
    color = "Faculty type"
  ) +
  theme_minimal()
```

The line plot shows how the percentage of different types of instructional staff changed over time. The improved plot is easier to understand than the original because the years are on the x-axis, percentages are on the y-axis, and each faculty type is shown with a separate colored line.

### Exercise 2

In my opinion, to make the story clearer, I would emphasize Part-Time Faculty because this group appears to increase over time compared with some other faculty types. I would use a stronger color or thicker line for Part-Time Faculty and make the other lines more muted? I think this would help the viewer to focus more on the main trend instead of trying to compare every line equally.

```{r Improve the plot by highlighting Part-Time Faculty}
staff_long %>%
  mutate(
    highlight = if_else(faculty_type == "Part-Time Faculty", 
                        "Part-Time Faculty", 
                        "Other faculty types")
  ) %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type
  )) +
  geom_line(aes(color = highlight, linewidth = highlight)) +
  scale_linewidth_manual(values = c("Part-Time Faculty" = 1.4,
                                    "Other faculty types" = 0.6)) +
  labs(
    title = "Part-Time Faculty Became a Larger Share of Instructional Staff",
    subtitle = "Instructional staff employment trends from 1975 to 2011",
    x = "Year",
    y = "Percentage of instructional staff",
    color = "Group",
    linewidth = "Group"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```

### Fisheries

### Exercise 3
The original fisheries plots are a bit difficult to understand as they stand. The 3D pie charts are hard to compare because the three-dimensional effect distorts the size of the slices. There are alos too many countries shown at once, which makes the visualization cluttered. To improve the visualization, I decided to use a horizontal bar chart. I will only show the top countries by total fisheries production so the plot is not overcrowded.

```{r load data for fisheries}
fisheries <- read_csv("data/fisheries.csv")
```

```{r Select the top 20 countries by total production}
top_countries <- fisheries %>%
  slice_max(total, n = 20) %>%
  pull(country)
```

```{r Change fisheries data from wide to long format}
fisheries_long <- fisheries %>%
  filter(country %in% top_countries) %>%
  pivot_longer(
    cols = c(capture, aquaculture),
    names_to = "type",
    values_to = "tons"
  ) %>%
  mutate(
    type = recode(type,
                  "capture" = "Capture fishing",
                  "aquaculture" = "Aquaculture")
  )

fisheries_long
```

```{r Make an improved fisheries bar chart visualization}
fisheries_long %>%
  ggplot(aes(
    x = tons,
    y = fct_reorder(country, tons, .fun = sum),
    fill = type
  )) +
  geom_col() +
  labs(
    title = "Top 20 Countries by Fisheries Production",
    subtitle = "Production is separated into capture fishing and aquaculture",
    x = "Production in tons",
    y = "Country",
    fill = "Production type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```

I think this improved plot is easier to read because it avoids 3D pie charts, limits the number of countries, and uses a horizontal layout so country names are readable. It also makes the distinction between capture fishing and aquaculture clearer. 


### Stretch Yourself with Smokers in Whickham
```{r Load the Whickham data and glimpse}
library(tidyverse)
library(mosaicData)
data(Whickham)
head(Whickham)
glimpse(Whickham)
```

### Exercise 4

These data come from an observational study. Participants were not randomly assigned to smoke or not smoke. Instead, their smoking status was observed, and their health outcome was recorded later.

### Exercise 5

```{r Count the number of observations}
nrow(Whickham)
```

There are `r nrow(Whickham)` observations in the dataset. Each observation represents one participant in the Whickham study.

### Exercise 6
```{r Show variable names and types}
glimpse(Whickham)
```

There are three variables in the dataset. 'Outcome', categorical variable showing whether the participant was alive or dead; 'smoker', categorical variable showing whether the participant smoked; and 'age', quantitative variable showing the participant’s age. 

```{r visualize each outcome}
# Visualize health outcome
ggplot(Whickham, aes(x = outcome, fill = outcome)) +
  geom_bar() +
  labs(
    title = "Distribution of Health Outcomes",
    x = "Outcome",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Visualize smoking status
ggplot(Whickham, aes(x = smoker, fill = smoker)) +
  geom_bar() +
  labs(
    title = "Distribution of Smoking Status",
    x = "Smoker",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Visualize age
ggplot(Whickham, aes(x = age)) +
  geom_histogram(binwidth = 5, color = "white") +
  labs(
    title = "Distribution of Age",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()
```

